Ecosystem size and complexity dictate riverine biodiversity

### 2021-01-25 manuscript.Rmd
minor updates

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity[@rinaldoEvolutionSelectionRiver2014; @rodriguez-iturbeFractalRiverBasins2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few[@rodriguez-iturbeFractalRiverBasins2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike[@rodriguez-iturbeFractalRiverBasins2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms[@grantLivingBranchesPopulation2007; @rinaldoRiverNetworksEcological2018;@tonkinRoleDispersalRiver2018]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure[@faganConnectivityFragmentationExtinction2002; @grantUseMultipleDispersal2010; @teruiMetapopulationStabilityBranching2018], and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems[@altermattDiversityRiverineMetacommunities2013; @carraraComplexInteractionDendritic2014; @carraraDendriticConnectivityControls2012; @teruiMetapopulationStabilityBranching2018]. However, most research has explored the consequences of branching complexity for local community structure[@altermattRiverNetworkProperties2013; @brownDendriticNetworkStructure2010] or has relied solely on theoretical arguments with limited replications of branching architecture[@holtRoleBranchingMaintenance2018]. At present, we lack a comprehensive evaluation of how branching complexity scales biodiversity at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where the accelerated species loss threatens the delivery of ecosystem services[@dudgeonFreshwaterBiodiversityImportance2006a].

Here, we hypothesize that ecosystem size and complexity dictate biodiversity patterns in rivers, a prime example of complex branching ecosystems. The recurrent merging of diverse streams and rivers produces larger, ecologically distinct channels, ultimately forming a fluvial network with fractal branching patterns[@rodriguez-iturbeFractalRiverBasins2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014], should control the ecosystem's habitat heterogeneity (habitat diversity per unit area). Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity). We predict that ecosystem size and branching complexity increase watershed-scale species richness ($\gamma$ diversity) by enhancing local species richness ($\alpha$ diversity) and/or spatial turnover of species composition ($\beta$ diversity) under different ecological scenarios. The present study combines theory and empirical analysis of extensive community data from two different regions of the globe to provide crucial insights into how ecological communities are structured in complex branching networks.

First, we theoretically analyzed the influences of ecosystem size and branching complexity on $\gamma$ diversity. We depicted branching ecosystems as a spatial network of connected habitat patches or local communities[@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014] (**Figure 1**) where two factors determine mean environmental conditions at each habitat patch: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution and propagate downstream with local environmental noise (i.e., a spatial autoregressive process with white noise). These environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see Methods). Therefore, our network-generation procedure resembles natural processes of how branching river networks create diverse habitats in a metacommunity. We then simulated metacommunity dynamics in the theoretical branching networks using a general metacommunity framework[@thompsonProcessbasedMetacommunityFramework2020]. In this model, 50 species with different abiotic niches (optimum and width) disperse along a network and compete for resources with varied strengths in temporally dynamic environments (see Methods for model details).

We considered `r n_scenario` simulation scenarios, which comprised a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`), both of which are defined as standard deviations of normal distributions. When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to minimal influences of branching on habitat heterogeneity. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity withing a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel), dispersal probability, and the maximum value of interspecific competition strength (see Methods). These simulation scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (Supplementary Information). Under each simulation scenario, we simulated 1000 time steps of metacommunity dynamics in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

@@

There may be confounding factors that are not incorporated into our statistical models. However, we are confident in our statistical inference because the observed patterns were fairly consistent with our theoretical predictions that are free from any confounding effects. Further, our fitted models generally showed high R^2^ values ($\alpha$ diversity: `r round(summary(fit$alpha$model)$r.squared, 2)`; $\beta$ diversity, `r round(summary(fit$beta$model)$r.squared, 2)`; $\gamma$ diversity `r round(summary(fit$gamma$model)$r.squared, 2)`) with no signs of spatial autocorrelation in residuals. Therefore, it is very unlikely that spurious correlations are responsible for our results.

@@

We depicted branching ecosystems as a bifurcating spatial network of connected habitat patches[@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014] (**Figure 1**). Habitat patches, or local communities, can be either non-branching or branching river sections with a unit length $l$, which defines the scale of local species interactions. Two parameters determine the geometric properties of simulated branching networks: the number of habitat patches $N_p$ (ecosystem size) and branching probability $P_b$ (ecosystem complexity). Each habitat patch is assigned to be a branching patch (including upstream terminals) with probability $P_b$ or a non-branching patch with probability $1-P_b$. In this framework, an individual branch is a consecutive series of non-branching patches terminated at a branching patch; therefore, the number of habitat patches in branch $q$, $n_{p,q}$, is a realization of a random variable drawn from a geometric distribution $n_{p,q} \sim Ge(P_{b})$[@teruiMetapopulationStabilityBranching2018] but conditional on $\sum_{q=1}^{N_b}n_{p,q} = N_p$ ($N_b$ is the number of branches in a network). The number of branches $N_b$ is identical to the number of branching habitat patches in a network, or the number of 'success' out of $N_p$ trials. This leads to an assumption that $N_b$ is a realization of a random variable drawn from a binomial distribution as $N_b \sim Binomial(N_p, P_b)$. This framework has two merits. First, it reflects observed patterns of branch length distribution, which is known to follow an exponential distribution (a continuous version of a geometric distribution)[@rodriguez-iturbeFractalRiverBasins2001]. Second, it preserves the fractal nature of branching patterns[@peckhamReformulationHortonLaws1999].

>>>>

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity[@rinaldoEvolutionSelectionRiver2014; @rodriguez-iturbeFractalRiverBasins2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few[@rodriguez-iturbeFractalRiverBasins2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike[@rodriguez-iturbeFractalRiverBasins2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms[@grantLivingBranchesPopulation2007; @rinaldoRiverNetworksEcological2018;@tonkinRoleDispersalRiver2018]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure[@faganConnectivityFragmentationExtinction2002; @grantUseMultipleDispersal2010; @teruiMetapopulationStabilityBranching2018], and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems[@altermattDiversityRiverineMetacommunities2013; @carraraComplexInteractionDendritic2014; @carraraDendriticConnectivityControls2012; @teruiMetapopulationStabilityBranching2018]. However, most research has explored the consequences of branching complexity for local community structure[@altermattRiverNetworkProperties2013; @brownDendriticNetworkStructure2010] or has relied solely on theoretical arguments with limited replications of branching architecture[@holtRoleBranchingMaintenance2018]. At present, we lack a comprehensive evaluation of how branching complexity scales biodiversity at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where the accelerated species loss threatens the delivery of ecosystem services[@dudgeonFreshwaterBiodiversityImportance2006].

Here, we hypothesize that ecosystem size and complexity dictate biodiversity patterns in rivers, a prime example of complex branching ecosystems. The recurrent merging of diverse streams and rivers produces larger, ecologically distinct channels, ultimately forming a fluvial network with fractal branching patterns[@rodriguez-iturbeFractalRiverBasins2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014a], should control the ecosystem's habitat heterogeneity (habitat diversity per unit area). Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity). We predict that ecosystem size and branching complexity increase watershed-scale species richness ($\gamma$ diversity) by enhancing local species richness ($\alpha$ diversity) and/or spatial turnover of species composition ($\beta$ diversity) under different ecological scenarios. The present study combines theory and empirical analysis of extensive community data from two different regions of the globe to provide crucial insights into how ecological communities are structured in complex branching networks.

First, we theoretically analyzed the influences of ecosystem size and branching complexity on $\gamma$ diversity. We depicted branching ecosystems as a spatial network of connected habitat patches or local communities[@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014a] (**Figure 1**) where two factors determine mean environmental conditions at each habitat patch: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution and propagate downstream with local environmental noise (i.e., a spatial autoregressive process with white noise). These environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see Methods). Therefore, our network-generation procedure resembles natural processes of how branching river networks create diverse habitats in a metacommunity. We then simulated metacommunity dynamics in the theoretical branching networks using a general metacommunity framework[@thompsonProcessbasedMetacommunityFramework2020]. In this model, 50 species with different abiotic niches (optimum and width) disperse along a network and compete for resources with varied strengths in temporally dynamic environments (see Methods for model details).

Given the results of extensive sensitivity analysis (**Supplementaly Information**), we considered `r n_scenario` simulation scenarios, which comprised a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`), both of which are defined as standard deviations of normal distributions. When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation. This landscape scenario should reproduce natural patterns of habitat heterogeneity, in which environmental conditions differ greatly among tributaries [@chezikRiverNetworksDampen2017; @mooreEmergentStabilityLarge2015; @petersonModellingDendriticEcological2013]. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to minimal influences of branching on habitat heterogeneity. This scenario may reflect human-modified landscapes where the physical or biological distinctiveness of tributaries has been compromised due to human activities[@mooreSynchronizationPortfolioPerformance2010; @poffHomogenizationRegionalRiver2007; @teruiMetapopulationStabilityBranching2018]. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity within a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel), dispersal probability, and the maximum value of interspecific competition strength (see Methods). These simulation scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (Supplementary Information). Under each simulation scenario, we simulated 1000 time steps of metacommunity dynamics in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

@@

There may be confounding factors that are not incorporated into our statistical models. However, we are confident in our statistical inference because the observed patterns were fairly consistent with our theoretical predictions that are free from any confounding effects. Further, our fitted models generally showed high R^2^ values ($\alpha$ diversity: `r round(summary(fit$alpha$model)$r.squared, 2)`; $\beta$ diversity, `r round(summary(fit$beta$model)$r.squared, 2)`; $\gamma$ diversity `r round(summary(fit$gamma$model)$r.squared, 2)`). Therefore, it is very unlikely that spurious correlations are responsible for our results.

@@

We depicted branching ecosystems as a bifurcating spatial network of connected habitat patches[@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014a] (**Figure 1**). Habitat patches, or local communities, can be either non-branching or branching river sections with a unit length $l$, which defines the scale of local species interactions. Two parameters determine the geometric properties of simulated branching networks: the number of habitat patches $N_p$ (ecosystem size) and branching probability $P_b$ (ecosystem complexity). Each habitat patch is assigned to be a branching patch (including upstream terminals) with probability $P_b$ or a non-branching patch with probability $1-P_b$. In this framework, an individual branch is a consecutive series of non-branching patches terminated at a branching patch; therefore, the number of habitat patches in branch $q$, $n_{p,q}$, is a realization of a random variable drawn from a geometric distribution $n_{p,q} \sim Ge(P_{b})$[@teruiMetapopulationStabilityBranching2018] but conditional on $\sum_{q=1}^{N_b}n_{p,q} = N_p$ ($N_b$ is the number of branches in a network). The number of branches $N_b$ is identical to the number of branching habitat patches in a network, or the number of 'success' out of $N_p$ trials. This leads to an assumption that $N_b$ is a realization of a random variable drawn from a binomial distribution as $N_b \sim Binomial(N_p, P_b)$. This framework has two merits. First, it reflects observed patterns of branch length distribution, which is known to follow an exponential distribution (a continuous version of a geometric distribution)[@rodriguez-iturbeFractalRiverBasins2001]. Second, it preserves the fractal nature of branching patterns[@peckhamReformulationHortonLaws1999].
